---
---
title: "RNA-seq analysis with RNASeq123 workflow"
author: "Kate, Jessica, Kardam"
bibliography: references.bib
format:
  gfm:
    output-file: README.md
    execute:
      echo: false
      eval: true
  revealjs:
    slide-level: 4
    execute:
      echo: false
      eval: true
---
---

























































## Differential Gene Analysis:
- To determine which genes are expressed at different levels between three cell population profiled.
- linear model fitting (assuming normally distributed data)
- Intercept act as anchor point for comparision against baseline. 
- Without intercept, absolute expression levels can be determined.

```{r}
design <- model.matrix(~0+group+lane)
colnames(design) <- gsub("group", "", colnames(design))
design
```




:::{.content-visible when-format="revealjs"}

- "makecontrast" function from limma package allows user to draw contrax matrix for pairwise comparisions.

- limma provides duplicateCorrelation to deal with technical replicates.

- limma provides flexibility to perform interactive studies for ex. testing drug reaction in Lane 6 and Lane 8.

```{r}
contr.matrix <- makeContrasts(
   BasalvsLP = Basal-LP, 
   BasalvsML = Basal - ML, 
   LPvsML = LP - ML, 
   levels = colnames(design))
contr.matrix
```


## Heteroscedascity and Voom weights:

- Homoscedascity: In linear models, mean-variance relationship is assumed to be linear. Mean and variance changes equally. Exactly opposite is 'Heteroscedascity'. 

- In RNA-Seq count data, the Negative Binomial distribution assumes quadratic mean-variation relationship. 

- Overdispersion observed due to large difference between house-keeping genes and highly expressed genes 

- In limma, linear modelling is carried out on log-CPM values.

- "voom" function act as link to bridge limma(built for microarrays) to expression data. It addresses scale problem (calculates into CPM) and "normalize.method" argument to normalize library size and estimates mean-variance relationship (non-linear - overdispersion) 

- "Voom weights" fixes the noise related to genes whereas Voomqualityweight addresses the noise generated from inter-sample variation.


```{r}
eval: FALSE
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
v
```


:::{.content-visible when-format="revealjs"}

```{r}
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```

Means (x-axis) and variances (y-axis) of each gene are plotted to show the dependence between the two before voom is applied to the data (left panel) and how the trend is removed after voom precision weights are applied to the data (right panel)

## Examining Differentially Expressed Genes:
Summary of Differentially expressed gene:
```{r}
summary(decideTests(efit))

```

Treat method - Method can be applied for stricter definition on significance based on t-statistics. This allows user to define a log-FC threshold. 

```{r}
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)
```


:::{.content-visible when-format="revealjs"}

Extracting DE genes:

```{r}
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)

de.common <- which(dt[,1]!=0 & dt[,2]!=0)
length(de.common)

head(tfit$genes$SYMBOL[de.common], n=20)

```

 O - represents genes with levels unchanged
 1 - Up regulated genes
-1 - Down regulated genes

:::{.content-visible when-format="revealjs"}

Venn diagram showing the number of genes DE in the comparison between basal versus LP only (left), basal versus ML only (right), and the number of genes that are DE in both comparisons (center)

The number of genes that are not DE in either comparison are marked in the bottom-right.

```{r}
vennDiagram(dt[,1:2], circle.col=c("turquoise", "salmon"))

```


## Examining individual DE genes from top to bottom:

- "topTreat"/ "topTable" - from toptreat or eBayes to list top DE genes.
 
- toptreat arranges DE genes chronologically in increasing order using log-FC, average log-CPM, moderated t-statistics, raw and adjusted p-value for each gene

- n=Inf -> all genes

```{r}
basal.vs.lp <- topTreat(tfit, coef=1, n=Inf)
basal.vs.ml <- topTreat(tfit, coef=2, n=Inf) 
head(basal.vs.lp)
```


## Graphical representation of DE gene results:

- Mean-difference plots are ideally utlized to display log-FC from linear model fit against log-CPM values using "plotMD" function.

- Glimma package offer interactive interface for MDplot by "glMDPlot".

- Glimma option allows brower viewing option - convenient for including them as linked files from an Rmarkdown.

```{r}
plotMD(tfit, column=1, status=dt[,1], main=colnames(tfit)[1], 
       xlim=c(-8,13))
```


:::{.content-visible when-format="revealjs"}

MDplot - allows user to visualize all genes or individual genes.

Heatmaps - Ideal option to visualize a subset of genes.

"heatmap.2"  from gplots package allows user to create heatmap for genes subset analysis.

# library(gplots)
# basal.vs.lp.topgenes <- basal.vs.lp$ENTREZID[1:100]
# i <- which(v$genes$ENTREZID %in% basal.vs.lp.topgenes)
# mycol <- colorpanel(1000,"blue","white","red")
# heatmap.2(lcpm[i,],   scale="row",
#  labRow=v$genes$SYMBOL[i], labCol=group, 
#  col=mycol, trace="none", density.info="none", 
# margin=c(8,6), lhei=c(2,10), dendrogram="column")

:::{.content-visible when-format="revealjs"}
```{r}
library(gplots)
basal.vs.lp.topgenes <- basal.vs.lp$ENTREZID[1:100]
i <- which(v$genes$ENTREZID %in% basal.vs.lp.topgenes)
mycol <- colorpanel(1000,"blue","white","red")
heatmap.2(lcpm[i,], scale="row",
   labRow=v$genes$SYMBOL[i], labCol=group, 
   col=mycol, trace="none", density.info="none", 
   margin=c(8,6), lhei=c(2,10), dendrogram="column")
```