---
title: "RNA-seq analysis with RNASeq123 workflow"
author: "Kate, Jessica, Kardam"
bibliography: References- BIFX 551 BioConductor Project.bib
format:
  gfm:
    output-file: README.md
  revealjs:
    slide-level: 4
    theme: [default, simple]
execute:
  echo: false
  output: false
  warning: false
  message: false
---

Load needed packages:


```{r}

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager", repos = "https://cloud.r-project.org")
}

if (!requireNamespace("RNAseq123", quietly = TRUE)) {
  BiocManager::install("RNAseq123", ask = FALSE, update = FALSE)
}

if (!requireNamespace("gplots", quietly = TRUE)) {
  install.packages("gplots", repos = "https://cloud.r-project.org")
}

library(limma)
library(Glimma)
library(edgeR)
library(Mus.musculus)
library(gplots)
library(RNAseq123)
```




Load Dependencies for flowchart

```{r}
if (!requireNamespace("ggflowchart", quietly = TRUE)) {
   install.packages("ggflowchart", repos = "https://cloud.r-project.org")
 }
require("ggflowchart")
library(ggflowchart)
library(tibble)
```


## Introduction

RNAseq123 Workflow:

```{r}
# Define steps
steps <- tibble(
    from = c("Set-up", "Data Packaging", "Data pre-processing", "Differential expression analysis"),
    to   = c("Data Packaging", "Data pre-processing", "Differential expression analysis", "Gene set testing with camera")
  
)

# Draw flowchart
ggflowchart(steps, colour = "green", family = "mono", text_size = 4, text_colour = "purple", arrow_colour = "orange", x_nudge = 0.35, y_nudge = 0.35)
```


## Experimental Data

“A pooled shRNA screen for regulators of primary mammary stem and progenitor cells identifies roles for Asap1 and Prox1”

This experiment used RNA-seq-based expression profiling in mouse mammary stem cell (MaSC)-enriched basal cells to look for candidate regulatory genes.

Mammary stem cells are often used in RNA-seq analyses as they are highly proliferative and have a high capacity for differentiation.

Regulatory genes produce proteins that turn expression on or off and are therefore important target genes for studying cancer as well as other developmental processes.

Using retroviral aided knockdown experiments and RNA-seq, the researchers were able to propose that the Asap1 gene is a negative regulator. 

The gene count file used in this analysis was created by aligning reads with the mouse reference genome using the align function (Rsubread) followed by gene-level summarization using featureCounts.

## Data Packaging 

- We started with raw RNA-seq count files from GEO accession GSE63310 and selected 9 samples
representing Basal, LP, and ML mammary cell populations.

- Using edgeR, we combined the individual count files into a single DGEList object, then added sample metadata such as biological group and sequencing lane. We also used Mus.musculus to attach gene annotations, including gene symbols and chromosome information.

::: {.content-visible when-format="revealjs"}

## Data Packaging 

:::

Data packaging converts separate raw files into one structured, analysis-ready
dataset.

```{r}
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE63310&format=file"
utils::download.file(url, destfile="GSE63310_RAW.tar", mode="wb") 
utils::untar("GSE63310_RAW.tar", exdir = ".")
files <- c("GSM1545535_10_6_5_11.txt", "GSM1545536_9_6_5_11.txt", "GSM1545538_purep53.txt",
  "GSM1545539_JMS8-2.txt", "GSM1545540_JMS8-3.txt", "GSM1545541_JMS8-4.txt",
  "GSM1545542_JMS8-5.txt", "GSM1545544_JMS9-P7c.txt", "GSM1545545_JMS9-P8c.txt")
for(i in paste(files, ".gz", sep=""))
  R.utils::gunzip(i, overwrite=TRUE)
```

## Data pre-processing

- Before comparing gene expression across groups, we transformed raw counts to CPM and log-CPM, filtered
out genes with very low expression, and normalized the libraries using TMM normalization with edgeR.

- We then used MDS plots to check whether samples clustered by biology rather than technical artifacts.

- This preprocessing step reduces noise, improves comparability across samples, and helps confirm that the data are ready for differential expression testing.

::: {.content-visible when-format="revealjs"}

## Data pre-processing

:::

Preprocessing removes uninformative genes, corrects library-size effects, and checks
overall sample quality before modeling.

```{r}
x <- readDGE(files, columns=c(1,3))
class(x)

dim(x)
```

```{r}
samplenames <- substring(colnames(x), 12, nchar(colnames(x)))
samplenames
```

```{r}
colnames(x) <- samplenames
group <- as.factor(c("LP", "ML", "Basal", "Basal", "ML", "LP", 
                     "Basal", "ML", "LP"))
x$samples$group <- group
lane <- as.factor(rep(c("L004","L006","L008"), c(3,4,2)))
x$samples$lane <- lane
x$samples
```


```{r}
geneid <- rownames(x)
genes <- select(Mus.musculus, keys=geneid, columns=c("SYMBOL", "TXCHROM"), 
                keytype="ENTREZID")
head(genes)
```


```{r}
genes <- genes[!duplicated(genes$ENTREZID),]
```

```{r}
x$genes <- genes
x
```

```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)

summary(lcpm)
```

```{r}
table(rowSums(x$counts==0)==9)
```

```{r}
keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```


```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
```

```{r}
x2 <- x
x2$samples$norm.factors <- 1
x2$counts[,1] <- ceiling(x2$counts[,1]*0.05)
x2$counts[,2] <- x2$counts[,2]*5
```

```{r}
par(mfrow=c(1,2))
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")
x2 <- calcNormFactors(x2)  
x2$samples$norm.factors
```

```{r}
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

```{r}
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
col.lane <- lane
levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
col.lane <- as.character(col.lane)
plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")
plotMDS(lcpm, labels=lane, col=col.lane, dim=c(3,4))
title(main="B. Sequencing lanes")
```

```{r}
glMDSPlot(lcpm, labels=paste(group, lane, sep="_"), 
          groups=x$samples[,c(2,5)], launch=FALSE)
```







## Differential Gene Analysis:
- To determine which genes are expressed at different levels between three cell population profiled.
- linear model fitting (assuming normally distributed data)
- Intercept act as anchor point for comparision against baseline. 
- Without intercept, absolute expression levels can be determined.

```{r}
design <- model.matrix(~0+group+lane)
colnames(design) <- gsub("group", "", colnames(design))
design
```




:::{.content-visible when-format="revealjs"}

- "makecontrast" function from limma package allows user to draw contrax matrix for pairwise comparisions.

- limma provides duplicateCorrelation to deal with technical replicates.

- limma provides flexibility to perform interactive studies for ex. testing drug reaction in Lane 6 and Lane 8.

```{r}
contr.matrix <- makeContrasts(
   BasalvsLP = Basal-LP, 
   BasalvsML = Basal - ML, 
   LPvsML = LP - ML, 
   levels = colnames(design))
contr.matrix
```
:::


## Heteroscedascity and Voom weights:

- Homoscedascity: In linear models, mean-variance relationship is assumed to be linear. Mean and variance changes equally. Exactly opposite is 'Heteroscedascity'. 

- In RNA-Seq count data, the Negative Binomial distribution assumes quadratic mean-variation relationship. 

- Overdispersion observed due to large difference between house-keeping genes and highly expressed genes 

- In limma, linear modelling is carried out on log-CPM values.

- "voom" function act as link to bridge limma(built for microarrays) to expression data. It addresses scale problem (calculates into CPM) and "normalize.method" argument to normalize library size and estimates mean-variance relationship (non-linear - overdispersion) 

- "Voom weights" fixes the noise related to genes whereas Voomqualityweight addresses the noise generated from inter-sample variation.


```{r}
#eval: FALSE
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
v
```


:::{.content-visible when-format="revealjs"}

```{r}
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```
:::

Means (x-axis) and variances (y-axis) of each gene are plotted to show the dependence between the two before voom is applied to the data (left panel) and how the trend is removed after voom precision weights are applied to the data (right panel)

## Examining Differentially Expressed Genes:
Summary of Differentially expressed gene:
```{r}
summary(decideTests(efit))

```

Treat method - Method can be applied for stricter definition on significance based on t-statistics. This allows user to define a log-FC threshold. 

```{r}
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)
```


:::{.content-visible when-format="revealjs"}

Extracting DE genes:

```{r}
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)

de.common <- which(dt[,1]!=0 & dt[,2]!=0)
length(de.common)

head(tfit$genes$SYMBOL[de.common], n=20)

```

 O - represents genes with levels unchanged
 1 - Up regulated genes
-1 - Down regulated genes
:::

:::{.content-visible when-format="revealjs"}

Venn diagram showing the number of genes DE in the comparison between basal versus LP only (left), basal versus ML only (right), and the number of genes that are DE in both comparisons (center)

The number of genes that are not DE in either comparison are marked in the bottom-right.

```{r}
vennDiagram(dt[,1:2], circle.col=c("turquoise", "salmon"))

```

:::

## Examining individual DE genes from top to bottom:

- "topTreat"/ "topTable" - from toptreat or eBayes to list top DE genes.
 
- toptreat arranges DE genes chronologically in increasing order using log-FC, average log-CPM, moderated t-statistics, raw and adjusted p-value for each gene

- n=Inf -> all genes

```{r}
basal.vs.lp <- topTreat(tfit, coef=1, n=Inf)
basal.vs.ml <- topTreat(tfit, coef=2, n=Inf) 
head(basal.vs.lp)
```


## Graphical representation of DE gene results:

- Mean-difference plots are ideally utlized to display log-FC from linear model fit against log-CPM values using "plotMD" function.

- Glimma package offer interactive interface for MDplot by "glMDPlot".

- Glimma option allows brower viewing option - convenient for including them as linked files from an Rmarkdown.

```{r}
plotMD(tfit, column=1, status=dt[,1], main=colnames(tfit)[1], 
       xlim=c(-8,13))
```


:::{.content-visible when-format="revealjs"}

MDplot - allows user to visualize all genes or individual genes.

Heatmaps - Ideal option to visualize a subset of genes.

"heatmap.2"  from gplots package allows user to create heatmap for genes subset analysis.

# library(gplots)
# basal.vs.lp.topgenes <- basal.vs.lp$ENTREZID[1:100]
# i <- which(v$genes$ENTREZID %in% basal.vs.lp.topgenes)
# mycol <- colorpanel(1000,"blue","white","red")
# heatmap.2(lcpm[i,],   scale="row",
#  labRow=v$genes$SYMBOL[i], labCol=group, 
#  col=mycol, trace="none", density.info="none", 
# margin=c(8,6), lhei=c(2,10), dendrogram="column")

:::

:::{.content-visible when-format="revealjs"}
```{r, fig.width=10, fig.height=14}
library(gplots)
basal.vs.lp.topgenes <- basal.vs.lp$ENTREZID[1:100]
i <- which(v$genes$ENTREZID %in% basal.vs.lp.topgenes)
mycol <- colorpanel(1000,"blue","white","red")
heatmap.2(lcpm[i,], scale="row",
   labRow=v$genes$SYMBOL[i], labCol=group, 
   col=mycol, trace="none", density.info="none", 
   margin=c(8,6), lhei=c(2,10), dendrogram="column")
```

:::