---
title: "RNA-seq analysis with RNAseq123 workflow"
author: "Kate, Jessica, Kardam"
bibliography: References- BIFX 551 BioConductor Project.bib
format:
  gfm:
    output-file: README.md
  revealjs:
    slide-level: 2
    theme: [default, simple]
execute:
  echo: false
  output: true
  warning: false
  message: false
---

```{r}
#load needed packages

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager", repos = "https://cloud.r-project.org")
}

if (!requireNamespace("RNAseq123", quietly = TRUE)) {
  BiocManager::install("RNAseq123", ask = FALSE, update = FALSE)
}

if (!requireNamespace("gplots", quietly = TRUE)) {
  install.packages("gplots", repos = "https://cloud.r-project.org")
}

library(limma)
library(Glimma)
library(edgeR)
library(Mus.musculus)
library(gplots)
library(RNAseq123)
```

```{r}
#load dependencies for flowchart 

if (!requireNamespace("ggflowchart", quietly = TRUE)) {
   install.packages("ggflowchart", repos = "https://cloud.r-project.org")
 }
require("ggflowchart")
library(ggflowchart)
library(tibble)
```


## Introduction

Goal: Demonstrate the RNAseq123 Bioconductor workflow

  Why this dataset?

   - Public mouse mammary cell RNA-seq counts from GSE63310
   - Includes biologically distinct groups: Basal, LP, and ML
   - Small, well-structured example for showing an end-to-end workflow

::: {.content-visible when-format="revealjs"}

## Introduction

:::

  What the workflow covers:

   - data packaging
   - filtering and normalization
   - quality assessment
   - differential expression analysis
   - visualization and interpretation

::: {.content-visible when-format="revealjs"}

## Introduction

:::

  This presentation is a worked example of how raw count data move through a standard
  Bioconductor RNA-seq workflow.

```{r}
 #| output: true
 #| fig-cap: "Overview of the RNA-seq workflow."
 #| fig-align: center
 #| out-width: 85%
# Define steps
steps <- tibble(
    from = c("Set-up", "Data Packaging", "Data pre-processing", "Differential expression analysis"),
    to   = c("Data Packaging", "Data pre-processing", "Differential expression analysis", "Gene set testing with camera")
  
)

# Draw flowchart
ggflowchart(steps, colour = "green", family = "mono", text_size = 4, text_colour = "purple", arrow_colour = "orange", x_nudge = 0.35, y_nudge = 0.35)
```


## Experimental Data

- “A pooled shRNA screen for regulators of primary mammary stem and progenitor cells identifies roles for Asap1 and Prox1”

- This experiment used RNA-seq-based expression profiling in mouse mammary stem cell (MaSC)-enriched basal cells to look for candidate regulatory genes.

- Mammary stem cells are often used in RNA-seq analyses as they are highly proliferative and have a high capacity for differentiation.

::: {.content-visible when-format="revealjs"}

## Experimental Data

:::

- Regulatory genes produce proteins that turn expression on or off and are therefore important target genes for studying cancer as well as other developmental processes.

- Using retroviral aided knockdown experiments and RNA-seq, the researchers were able to propose that the Asap1 gene is a negative regulator. 

- The gene count file used in this analysis was created by aligning reads with the mouse reference genome using the align function (Rsubread) followed by gene-level summarization using featureCounts.

## Data Packaging 

- We started with raw RNA-seq count files from GEO accession GSE63310 and selected 9 samples
representing Basal, LP, and ML mammary cell populations.

- Using edgeR, we combined the individual count files into a single DGEList object, then added sample metadata such as biological group and sequencing lane. We also used Mus.musculus to attach gene annotations, including gene symbols and chromosome information.

::: {.content-visible when-format="revealjs"}

## Data Packaging 

:::

Data packaging converts separate raw files into one structured, analysis-ready
dataset.

```{r}
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE63310&format=file"
utils::download.file(url, destfile="GSE63310_RAW.tar", mode="wb") 
utils::untar("GSE63310_RAW.tar", exdir = ".")
files <- c("GSM1545535_10_6_5_11.txt", "GSM1545536_9_6_5_11.txt", "GSM1545538_purep53.txt",
  "GSM1545539_JMS8-2.txt", "GSM1545540_JMS8-3.txt", "GSM1545541_JMS8-4.txt",
  "GSM1545542_JMS8-5.txt", "GSM1545544_JMS9-P7c.txt", "GSM1545545_JMS9-P8c.txt")
for(i in paste(files, ".gz", sep=""))
  R.utils::gunzip(i, overwrite=TRUE)
```

## Data Pre-Processing

- Before comparing gene expression across groups, we transformed raw counts to CPM and log-CPM, filtered
out genes with very low expression, and normalized the libraries using TMM normalization with edgeR.



```{r}
#| include: false
x <- readDGE(files, columns=c(1,3))
class(x)

dim(x)
```

```{r}
#| include: false
samplenames <- substring(colnames(x), 12, nchar(colnames(x)))
samplenames
```

```{r}
#| include: false
colnames(x) <- samplenames
group <- as.factor(c("LP", "ML", "Basal", "Basal", "ML", "LP", 
                     "Basal", "ML", "LP"))
x$samples$group <- group
lane <- as.factor(rep(c("L004","L006","L008"), c(3,4,2)))
x$samples$lane <- lane
x$samples
```


```{r}
#| include: false
geneid <- rownames(x)
genes <- select(Mus.musculus, keys=geneid, columns=c("SYMBOL", "TXCHROM"), 
                keytype="ENTREZID")
head(genes)
```


```{r}
#| include: false
genes <- genes[!duplicated(genes$ENTREZID),]
```

```{r}
#| include: false
x$genes <- genes
x
```

```{r}
#| include: false
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

```{r}
#| include: false
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)

summary(lcpm)
```

```{r}
#| include: false
table(rowSums(x$counts==0)==9)
```

```{r}
#| include: false
keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

```{r}
#| include: false
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```


```{r}
#| include: false
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
```

```{r}
#| include: false
x2 <- x
x2$samples$norm.factors <- 1
x2$counts[,1] <- ceiling(x2$counts[,1]*0.05)
x2$counts[,2] <- x2$counts[,2]*5
```

```{r}
 #| results: "hide"
 par(mfrow=c(1,2))
 
 lcpm <- cpm(x2, log=TRUE)
 boxplot(lcpm, las=2, col=col, main="")
 title(main="A. Example: Unnormalised data", ylab="Log-cpm")
 
 x2 <- calcNormFactors(x2)
 lcpm <- cpm(x2, log=TRUE)
 boxplot(lcpm, las=2, col=col, main="")
 title(main="B. Example: Normalised data", ylab="Log-cpm")
```

```{r}
#combined this chunk and the one below it in the code block above
#par(mfrow=c(1,2))
#lcpm <- cpm(x2, log=TRUE)
#boxplot(lcpm, las=2, col=col, main="")
#title(main="A. Example: Unnormalised data",ylab="Log-cpm")
#x2 <- calcNormFactors(x2)  
#x2$samples$norm.factors
```

```{r}
#lcpm <- cpm(x2, log=TRUE)
#boxplot(lcpm, las=2, col=col, main="")
#title(main="B. Example: Normalised data",ylab="Log-cpm")
```

```{r}
#| include: false
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
col.lane <- lane
levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
col.lane <- as.character(col.lane)
plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")
plotMDS(lcpm, labels=lane, col=col.lane, dim=c(3,4))
title(main="B. Sequencing lanes")
```

```{r}
#| include: false
glMDSPlot(lcpm, labels=paste(group, lane, sep="_"), 
          groups=x$samples[,c(2,5)], launch=FALSE)
```


::: {.content-visible when-format="revealjs"}

## Data Pre-Processing

:::

- We then used MDS plots to check whether samples clustered by biology rather than technical artifacts.

- This preprocessing step reduces noise, improves comparability across samples, and helps confirm that the data are ready for differential expression testing.

::: {.content-visible when-format="revealjs"}

## Data Pre-Processing

:::

Preprocessing removes uninformative genes, corrects library-size effects, and checks
overall sample quality before modeling.




