---
title: "RNA123_analysis_file"
format: html
editor: visual
authors: Kate Kruczynski, Jessica Joe, Kardam Kiranbhai Patel
---

Prepare Environment for analysis:

Install RNA123 package in RGui as recommended by BioConductor instructions.

Load needed packages:

```{r}
library(limma)
library(Glimma)
library(edgeR)
library(Mus.musculus)
```

Load in test case data:

Only selected files are grouped into "files" to create a smaller dataset only including the basal, ML, and LP samples from the experiment. The function **gunzip** from R.utils unzips the .gz files.

```{r}
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE63310&format=file"
utils::download.file(url, destfile="GSE63310_RAW.tar", mode="wb") 
utils::untar("GSE63310_RAW.tar", exdir = ".")
files <- c("GSM1545535_10_6_5_11.txt", "GSM1545536_9_6_5_11.txt", "GSM1545538_purep53.txt",
  "GSM1545539_JMS8-2.txt", "GSM1545540_JMS8-3.txt", "GSM1545541_JMS8-4.txt",
  "GSM1545542_JMS8-5.txt", "GSM1545544_JMS9-P7c.txt", "GSM1545545_JMS9-P8c.txt")
for(i in paste(files, ".gz", sep=""))
  R.utils::gunzip(i, overwrite=TRUE)
```

Each file represents the data from an individual sample. Looking at the first few lines of the first file we can see that it contains gene counts by EntrezID and also the length of each gene.

```{r}
read.delim(files[1], nrow=5)

```

The EdgeR package contains the function readDGE, which can take the individual files uploaded and create a count matrix.

```{r}
x <- readDGE(files, columns=c(1,3))
class(x)

dim(x)
```

The DGElist "x" that was created contains the count matrix and a dataframe called "samples" which contains experimental metadata that we want to associate with the counts.

```{r}
samplenames <- substring(colnames(x), 12, nchar(colnames(x)))
samplenames
```

```{r}
colnames(x) <- samplenames
group <- as.factor(c("LP", "ML", "Basal", "Basal", "ML", "LP", 
                     "Basal", "ML", "LP"))
x$samples$group <- group
lane <- as.factor(rep(c("L004","L006","L008"), c(3,4,2)))
x$samples$lane <- lane
x$samples
```

In order to annotate the genes in the counts matrix we can use the package Mus. musculus, which interfaces with Ensembl genome databases to pull gene symbols and chromosome information based on the Entrez IDs in our genes dataframe.

```{r}
geneid <- rownames(x)
genes <- select(Mus.musculus, keys=geneid, columns=c("SYMBOL", "TXCHROM"), 
                keytype="ENTREZID")
head(genes)
```

There are some duplicated gene IDs in our dataset due to genes mapping to multiple chromosomes. A simple, although not the most accurate, way to eliminate the duplicates is to only keep the first occurrence of each gene ID.

```{r}
genes <- genes[!duplicated(genes$ENTREZID),]
```

We can then add the cleaned dataframe back into the DGElist object x. Now we have a count matrix with associated experimental information and gene information, ready for further processing.

```{r}
x$genes <- genes
x
```

The raw counts need to be transformed to account for differences in sequencing depth between libraries. The EdgeR functions **cpm** and **lcpm** can be used to transform the raw counts into "counts per million" and "log2 counts per million", respectively.

```{r}
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
```

A count of zero for this data equates to log- CPM value of -4.51.

```{r}
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)

summary(lcpm)
```

Genes that are not expressed (or very lowly expressed) in any of the experimental conditions are not of interest in this analysis. To simplify all downstream analysis, the EdgeR function **filterByExpr** can be used to filter out all genes that have insignificant counts.

```{r}
table(rowSums(x$counts==0)==9)
```

```{r}
keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
```

The code below will produce a figure by which we can compare our raw data to our newly filtered and cleaned data.

```{r}
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
```

Due to the large number of variables and possible batch effects in sequencing experiments, all gene expression data should be normalized prior to further analysis. The edgeR function **calcNormFactors** can be used to normalize by the method of trimmed mean of M-values.

```{r}
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
```

The effect of the normalization is not large in this dataset as the distribution of log-CPM values are similar for the samples in this dataset. The code below will adjust the distributions to give a better visualization of the effects of a normalization.

```{r}
x2 <- x
x2$samples$norm.factors <- 1
x2$counts[,1] <- ceiling(x2$counts[,1]*0.05)
x2$counts[,2] <- x2$counts[,2]*5
```

```{r}
par(mfrow=c(1,2))
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")
x2 <- calcNormFactors(x2)  
x2$samples$norm.factors
```

```{r}
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```

An unsupervised clustering analysis can be performed to take a quick look at the magnitude of differential expression that we can expect to find in further analyses. It can also call out possible issues such as sample switches if any replicates or samples of the same grouping are far from each other. The limma function **plotMDS** can be used a create multi-dimensional scaling plots.

```{r}
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
col.lane <- lane
levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
col.lane <- as.character(col.lane)
plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")
plotMDS(lcpm, labels=lane, col=col.lane, dim=c(3,4))
title(main="B. Sequencing lanes")
```

Or, alternatively, the Glimma function **glMDSPlot** can be used to create an interactive plot in an html page.

```{r}
glMDSPlot(lcpm, labels=paste(group, lane, sep="_"), 
          groups=x$samples[,c(2,5)], launch=FALSE)
```

Differential expression analysis:

Contrast Matrix

The stats package function **model.matrix** is used to create a design matrix wherein categorical variables are transformed to numeric variables, so that the data can be fitted to regression models. Here the variables tested are sample group(cell type) and sequencing lane. The "\~0+group" removes the intercept from the first factor, group, but not from the second factor, lane.

```{r}
design <- model.matrix(~0+group+lane)
colnames(design) <- gsub("group", "", colnames(design))
design
```

The limma function **makeContrasts** can be used to make pair-wise comparisons between factors.

```{r}
contr.matrix <- makeContrasts(
   BasalvsLP = Basal-LP, 
   BasalvsML = Basal - ML, 
   LPvsML = LP - ML, 
   levels = colnames(design))
contr.matrix
```

Removing Heteroscedascity:

Heteroscedascity occurs when the variance (measure of error) is not the same across the independent variables.

```{r}
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
v
```

```{r}
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```

Differential expression genes summary:

The limma package contains several functions for linear modeling that fit a separate model to the expression values for each gene. The **decideTests** function can take the fitted model object and classify each gene as being significantly downregulated, upregulated, or not significant. A summary of these results is produced below.

```{r}
summary(decideTests(efit))
```

Some studies require that the log-fold-changes be above a certain value to be considered significant. The "treat" method can be used to specify this minimum value. In the table below one can see that this specification does place considerably more genes in the Not Significant categories.

```{r}
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)
```

The following code can be used to find genes that are DE in more than one group.The number of DE genes in common between groups can be easily summarized using **vennDiagram.**

```{r}
de.common <- which(dt[,1]!=0 & dt[,2]!=0)
length(de.common)
```

```{r}
head(tfit$genes$SYMBOL[de.common], n=20)
```

```{r}
vennDiagram(dt[,1:2], circle.col=c("turquoise", "salmon"))
```

The **write.fit** function will pull the results from the comparisons and write them to a file.

```{r}
write.fit(tfit, dt, file="results.txt")
```

Individual DE genes:

The **topTreat** (or **topTable** for results using eBayes statistics) function will show a list of the DE genes from smallest to largest adjusted p-value.

```{r}
basal.vs.lp <- topTreat(tfit, coef=1, n=Inf)
basal.vs.ml <- topTreat(tfit, coef=2, n=Inf)
head(basal.vs.lp)
```

```{r}
head(basal.vs.ml)
```

Graphics for DE results:

A mean-difference plot can be used to visually summarize the results for all genes. The **plotMD** function can display the log-FC values against the average log-CPM values. As with most plots, the Glimma package can also be used to create an interactive plot in a separate html page.

```{r}
plotMD(tfit, column=1, status=dt[,1], main=colnames(tfit)[1], 
       xlim=c(-8,13))
```

```{r}
glMDPlot(tfit, coef=1, status=dt, main=colnames(tfit)[1],
         side.main="ENTREZID", counts=lcpm, groups=group, launch=FALSE)
```

Heat maps are very useful for visualizing portions of DE data. The **heatmap.2** function from the gplots package is used here to create a heatmap of the top 100 DE genes from the basal vs LP contrast analysis.

```{r}
library(gplots)
basal.vs.lp.topgenes <- basal.vs.lp$ENTREZID[1:100]
i <- which(v$genes$ENTREZID %in% basal.vs.lp.topgenes)
mycol <- colorpanel(1000,"blue","white","red")
heatmap.2(lcpm[i,], scale="row",
   labRow=v$genes$SYMBOL[i], labCol=group, 
   col=mycol, trace="none", density.info="none", 
   margin=c(8,6), lhei=c(2,10), dendrogram="column")
```

Gene set testing with camera

```{r}
load(system.file("extdata", "mouse_c2_v5p1.rda", package = "RNAseq123"))
idx <- ids2indices(Mm.c2,id=rownames(v))
cam.BasalvsLP <- camera(v,idx,design,contrast=contr.matrix[,1])
head(cam.BasalvsLP,5)
```

```{r}
cam.BasalvsML <- camera(v,idx,design,contrast=contr.matrix[,2])
head(cam.BasalvsML,5)
```

```{r}
cam.LPvsML <- camera(v,idx,design,contrast=contr.matrix[,3])
head(cam.LPvsML,5)
```

```{r}
barcodeplot(efit$t[,3], index=idx$LIM_MAMMARY_LUMINAL_MATURE_UP, 
            index2=idx$LIM_MAMMARY_LUMINAL_MATURE_DN, main="LPvsML")
```
